The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2011-10-18 pkg/yuimadocs/inst/doc/JSS/article.Rnw
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@194 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% almost as usual
\author{Alexandre Brouste\\University of Le Mans \And 
        Masaaki Fukasawa\\University of Osaka\And
        Hideitsu Hino\\Waseda University\And
        Stefano M. Iacus\\University of Milan \AND 	
        Kengo Kamatani\\University of Tokyo \And
        Yuta Koike\\University of Tokyo\And
        Hiroki Masuda\\University of Kyushu \And
        Ryosuke Nomura\\University of Tokyo\AND
        Yasutaka Shimuzu\\University of Osaka \And
        Masayuki Uchida\\University of Osaka \And
        Nakahiro Yoshida\\University of Tokyo 
        }
\title{The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations}

@@

\subsection{The $\tt yuima.model$ class}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified. All multidimensional and eventually as parametric models.
\begin{itemize}
\item diffusions $\displaystyle  \de X_t=a(t,X_t)dt + b(t,X_t)\de W_t $, where $W_t$ is a standard Brownian motion;
\item fractional Gaussian noise, with $H$ the Hurst parameter
$$ \de X_t=a(t,X_t)dt + b(t,X_t)\de W_t^{H}; $$ 
\item diffusions with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(X_t)dt + b(X_t)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}.
\end{aligned}
$$
\end{itemize}

@@

\subsection{L\'evy processes}
Jump processes can be specified in different ways in mathematics and hence in \pkg{yuima} package. 
Let $Z_t$ be a  Compound Poisson Process (i.e. jumps size follow some distribution, like the Gaussian law and jumps occur at Poisson times).
Then it is possible to consider the following SDE which involves jumps
$$\de X_t =  a(X_t)dt + b(X_t)\de W_t + \de Z_t$$
In the next example we consider a compound Poisson process with intensity $\lambda=10$ with Gaussian jumps.
This model can be specified in \code{setModel} using the argument  \code{measure.type="CP"} 
A simple Ornstein-Uhlembeck process with Gaussian jumps 
$$\de X_t = -\theta X_t \de t + \sigma \de W_t + Z_t$$
is specified as
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod5 <- setModel(drift=c("-theta*x"), diffusion="sigma",
 jump.coeff="1", measure=list(intensity="10", df=list("dnorm(z, 0, 1)")),
 measure.type="CP", solve.variable="x")
set.seed(123)
X <- simulate(mod5, true.p=list(theta=1,sigma=3),sampling=setSampling(n=1000))
plot(X)
@

\noindent
Another possibility is to specify the jump component via the L\'evy measure. Without going into too much details, here is an example of specification of a simple Ornstein-Uhlembeck  process with IG (Inverse Gaussian) L\'evy measure 
$$\de X_t = -x dt + dZ_t$$
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod6 <- setModel(drift="-x", xinit=1, jump.coeff="1", 
  measure.type="code", measure=list(df="rIG(z, 1, 0.1)"))
set.seed(123)
X <- simulate(mod6, sampling=setSampling(Terminal=10, n=10000)) 
plot(X)
@

\subsection{Specification of generic models}
In general, the \pkg{yuima} package allows to specify a large family of models solutions to
$$\de X_t =a(X_t)\de t +  b(X_t)\de W_t + c(X_t)\de Z_t$$
using the following interface
<<echo=TRUE, print=FALSE,eval=FALSE>>=
setModel(drift, diffusion, hurst = 0.5, jump.coeff,
 measure, measure.type, state.variable = "x",
 jump.variable = "z", time.variable = "t", solve.variable, xinit)
@
The \pkg{yuima} package implements many multivariate Random Numbers Generators (RNG) which are needed to simulate L\'evy paths including
\code{rIG} (Inverse Gaussian), \code{rNIG} (Normal Inverse Gaussian), \code{rbgamma} (Bilateral Gamma), \code{rngamma} (Gamma) and \texttt{rstable} (Stable Laws).
Other user-defined RNG can be used freely.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<eval=TRUE,echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<eval=TRUE,echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{Small sample size}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^(1/3)$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and the value of $T$ is then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

\subsection{Change point analysis}
Consider a multidimensional stochastic differential equation of the form
$$
\de Y_t = b_t \de t + \sigma(X_t,\theta) \de W_t,\ \ t\in[0,T], 
$$
where $W_t$ a $r$-dimensional Wiener process and $b_t$ and $X_t$ are multidimensional processes and $\sigma$ is the diffusion coefficient (volatility) matrix.
When $Y=X$ the problem is a diffusion model.
The process $b_t$ may have jumps but should not explode and it is treated as a nuisance in this model.
The change-point problem for the volatility is formalized as follows
$$
Y_t=
\Bigg\{
\begin{array}{ll}
Y_0+\int_0^t b_s d s+\int_0^t \sigma(X_s,\theta_0^*) d W_s
& \mbox{ for } t\in[0,\tau^*)
\\
Y_{\tau^*}+\int_{\tau^*}^t b_s d s+\int_{\tau^*}^t \sigma(X_s,\theta_1^*) d W_s
& \mbox{ for } t\in[\tau^*,T]. 
\end{array}
$$
The change point $\tau^*$ instant is unknown and 
is  to be estimated, along with $\theta_0^*$ and $\theta_1^*$, from the observations sampled 
from the path of $(X,Y)$. The \pkg{yuima} implements the quasi-maximum likelihood approach as in \citet{iacyos09} described in the following.
Let  $\Delta_iY=Y_{t_i}-Y_{t_{i-1}}$ and 
define 
\begin{equation} 
\Phi_n(t;\theta_0,\theta_1)
=
\sum_{i=1}^{[nt{/T}]}G_i(\theta_0)+\sum_{i={[nt{/T}]+1}}^nG_i(\theta_1),
\label{eq:phiCH10}
\end{equation}
with 
\begin{equation} 
G_i(\theta) =
\log\det S(X_{t_{i-1}},\theta)
+ \Delta_n^{-1}(\Delta_iY)'S(X_{t_{i-1}},\theta)^{-1}(\Delta_iY). 
\label{eq:GiCH10}
\end{equation}
Suppose that there exists an estimator 
$\hat{\theta}_k$ for each $\theta_k$, $k=0,1$. 
In case $\theta_k^*$ are known, we define $\hat{\theta}_k$ 
just as $\hat{\theta}_k=\theta_k^*$. 
The change point estimator  of $\tau^*$ is
\begin{eqnarray*} 
\hat{\tau}_n&=&\arg\!\!\min\limits_{t\in[0,T]} 
\Phi_n(t;\hat{\theta}_0,\hat{\theta}_1).
\end{eqnarray*}

>>>>

%% almost as usual
\author{Alexandre Brouste\\University of Le Mans \And 
        Masaaki Fukasawa\\Osaka University\And
        Hideitsu Hino\\Waseda University\And
        Stefano M. Iacus\\University of Milan \AND 	
        Kengo Kamatani\\The University of Tokyo \And
        Yuta Koike\\The University of Tokyo\And
        Hiroki Masuda\\Kyushu University \And
        Ryosuke Nomura\\The University of Tokyo\AND
        Yasutaka Shimuzu\\Osaka University \And
        Masayuki Uchida\\Osaka University \And
        Nakahiro Yoshida\\The University of Tokyo 
        }
\title{The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations}

@@

\subsection{The $\tt yuima.model$ class}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified. All multidimensional and eventually as parametric models.
\begin{itemize}
\item diffusions $\displaystyle  \de X_t=a(t,X_t)dt + b(t,X_t)\de W_t $, where $W_t$ is a standard Brownian motion;
\item fractional Gaussian noise, with $H$ the Hurst parameter
$$ \de X_t=a(t,X_t)dt + b(t,X_t)\de W_t^{H}; $$ 
\item diffusions with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(X_t)\de t + b(X_t)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}.
\end{aligned}
$$
\end{itemize}

@@

\subsection{L\'evy processes}
Jump processes can be specified in different ways in mathematics and hence in \pkg{yuima} package. 
Let $Z_t$ be a  Compound Poisson Process (i.e. jumps size follow some distribution, like the Gaussian law and jumps occur at Poisson times).
Then it is possible to consider the following SDE which involves jumps
$$\de X_t =  a(X_t)\de t + b(X_t)\de W_t + \de Z_t$$
In the next example we consider a compound Poisson process with intensity $\lambda=10$ with Gaussian jumps.
This model can be specified in \code{setModel} using the argument  \code{measure.type="CP"} 
A simple Ornstein-Uhlembeck process with Gaussian jumps 
$$\de X_t = -\theta X_t \de t + \sigma \de W_t + \de Z_t$$
is specified as
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod5 <- setModel(drift=c("-theta*x"), diffusion="sigma",
 jump.coeff="1", measure=list(intensity="10", df=list("dnorm(z, 0, 1)")),
 measure.type="CP", solve.variable="x")
set.seed(123)
X <- simulate(mod5, true.p=list(theta=1,sigma=3),sampling=setSampling(n=1000))
plot(X)
@

\noindent
Another possibility is to specify the jump component via the L\'evy measure. Without going into too much details, here is an example of specification of a simple Ornstein-Uhlembeck  process with IG (Inverse Gaussian) L\'evy measure 
$$\de X_t = -x \de t + \de Z_t$$
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
mod6 <- setModel(drift="-x", xinit=1, jump.coeff="1", 
  measure.type="code", measure=list(df="rIG(z, 1, 0.1)"))
set.seed(123)
X <- simulate(mod6, sampling=setSampling(Terminal=10, n=10000)) 
plot(X)
@

\subsection{Specification of generic models}
In general, the \pkg{yuima} package allows to specify a large family of models solutions to
$$\de X_t =a(X_t)\de t +  b(X_t)\de W_t + c(X_t)\de Z_t$$
using the following interface
<<echo=TRUE, print=FALSE,eval=FALSE>>=
setModel(drift, diffusion, hurst = 0.5, jump.coeff,
 measure, measure.type, state.variable = "x",
 jump.variable = "z", time.variable = "t", solve.variable, xinit)
@
The \pkg{yuima} package implements many multivariate Random Numbers Generators (RNG) which are needed to simulate L\'evy paths including
\code{rIG} (Inverse Gaussian), \code{rNIG} (Normal Inverse Gaussian), \code{rbgamma} (Bilateral Gamma), \code{rngamma} (Normal Gamma) and \texttt{rstable} (Stable Laws).
Other user-defined RNG can be used freely.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<eval=TRUE,echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<eval=TRUE,echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{Small sample size}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^\frac13$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and the value of $T$ is then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

\subsection{Change point analysis}
Consider a multidimensional stochastic differential equation of the form
$$
\de Y_t = b_t \de t + \sigma(X_t,\theta) \de W_t,\ \ t\in[0,T], 
$$
where $W_t$ a $r$-dimensional Wiener process and $b_t$ and $X_t$ are multidimensional processes and $\sigma$ is the diffusion coefficient (volatility) matrix.
When $Y=X$ the problem is a diffusion model.
The process $b_t$ may have jumps but should not explode and it is treated as a nuisance in this model.
The change-point problem for the volatility is formalized as follows
$$
Y_t=
\Bigg\{
\begin{array}{ll}
Y_0+\int_0^t b_s \de s+\int_0^t \sigma(X_s,\theta_0^*) \de W_s
& \mbox{ for } t\in[0,\tau^*)
\\
Y_{\tau^*}+\int_{\tau^*}^t b_s \de s+\int_{\tau^*}^t \sigma(X_s,\theta_1^*) \de W_s
& \mbox{ for } t\in[\tau^*,T]. 
\end{array}
$$
The change point $\tau^*$ instant is unknown and 
is  to be estimated, along with $\theta_0^*$ and $\theta_1^*$, from the observations sampled 
from the path of $(X,Y)$. The \pkg{yuima} implements the quasi-maximum likelihood approach as in \citet{iacyos09} described in the following.
Let  $\Delta_iY=Y_{t_i}-Y_{t_{i-1}}$ and 
define 
\begin{equation} 
\Phi_n(t;\theta_0,\theta_1)
=
\sum_{i=1}^{[nt{/T}]}G_i(\theta_0)+\sum_{i={[nt{/T}]+1}}^nG_i(\theta_1),
\label{eq:phiCH10}
\end{equation}
with 
\begin{equation} 
G_i(\theta) =
\log\det S(X_{t_{i-1}},\theta)
+ \Delta_n^{-1}(\Delta_iY)'S(X_{t_{i-1}},\theta)^{-1}(\Delta_iY). 
\label{eq:GiCH10}
\end{equation}
Suppose that there exists an estimator 
$\hat{\theta}_k$ for each $\theta_k$, $k=0,1$. 
In case $\theta_k^*$ are known, we define $\hat{\theta}_k$ 
just as $\hat{\theta}_k=\theta_k^*$. 
The change point estimator  of $\tau^*$ is
\begin{eqnarray*} 
\hat{\tau}_n&=&\arg\!\!\min\limits_{t\in[0,T]} 
\Phi_n(t;\hat{\theta}_0,\hat{\theta}_1).
\end{eqnarray*}
